Importance of depth refugia for reef resilience and intervention on the Great Barrier Reef under future climate change

### 2025-10-13 paper/paper.qmd
Update format config
Added authors-block extension

Global Change Biology citation style

Fixed/updated bibtex references

---
title: Importance of depth refugia for reef resilience and intervention on the Great Barrier Reef under future climate change
author:
  - name: Benjamin Grier
    email: grierben777@gmail.com
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Townsville, Queensland, Australia
    attributes:
        corresponding: true
  - name: Takuya Iwanaga
    email: t.iwanaga@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Townsville, Queensland, Australia
  - name: Pedro Ribeiro De Almeida
    email: p.ribeirodealmeida@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Townsville, Queensland, Australia
  - name: Chinenye Ani
    email: c.ani@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Townsville, Queensland, Australia
  - name: Sam Matthews
    email: s.matthews@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Townsville, Queensland, Australia
abstract: |
    Coral reefs globally face unprecedented threats from climate change, with the Great
    Barrier Reef (GBR) experiencing cumulative stressors and increasingly severe declines
    in coral cover from thermal stress events. Understanding drivers behind reef resilience
    to climate impacts is critical for conservation planning and intervention strategies. A
    scenario-based population modelling approach was adopted with larval connectivity dynamics and
    environmental factors to assess coral reef resilience across the GBR using projected conditions under
    five Global Climate Models (GCMs). Projected coral
    cover was analysed for each reef's ability to maintain positive carbonate production budgets under
    future conditions, using coral cover as a proxy. Larval connectivity patterns did not correlate with
    increases in maintenance of positive carbonate budgets. Instead, reef depth emerged as the primary
    predictor, with deeper reefs benefitting from thermal refugia that reduce bleaching impacts. These
    findings suggest that depth is a tangible and pragmatic reef characteristic to consider in future
    intervention practices for coral reef restoration. These results have important implications for reef
    management, indicating that depth refugia should be prioritized for further consideration in
    conservation planning to maximize coral survival under continuing climate change.

keywords:
  - Coral reef
  - Climate change
  - Ecosystem resilience
  - Reef depth
  - Reef connectivity
date: last-modified
format:
  elsevier-docx:
    journal:
      name: Journal Name
      formatting: preprint
      cite-style: authoryear
bibliography: references.bib
engine: julia
crossref:
  custom:
  - kind: float
    reference-prefix: SFigure
    key: sfig
  - kind: float
    reference-prefix: STable
    key: stbl
editor:
  markdown:
    wrap: 92
    canonical: true
---

>>>>

---
title: Importance of depth refugia for reef resilience and intervention on the Great Barrier Reef under future climate change

authors:
  - name: Benjamin Grier
    email: grierben777@gmail.com
    affiliations:
      - ref: aims
    orcid: 0009-0005-1444-6198
    attributes:
      corresponding: true

  - name: Takuya Iwanaga
    email: t.iwanaga@aims.gov.au
    affiliations:
      - ref: aims
    orcid: 0000-0001-8173-0870

  - name: Pedro Ribeiro De Almeida
    email: p.ribeirodealmeida@aims.gov.au
    affiliations:
      - ref: aims
    orcid: 0009-0007-4814-3261

  - name: Chinenye Ani
    email: c.ani@aims.gov.au
    affiliations:
      - ref: aims
    orcid: 0000-0002-2112-0840

  - name: Sam Matthews
    email: s.matthews@aims.gov.au
    affiliations:
      - ref: aims
    orcid: 0000-0003-1936-1220

affiliations:
  - id: aims
    name: Australian Institute of Marine Science, Townsville, Queensland, Australia

abstract: |
    Coral reefs globally face unprecedented threats from climate change, with the Great
    Barrier Reef (GBR) experiencing cumulative stressors and increasingly severe declines
    in coral cover from thermal stress events. Understanding drivers behind reef resilience
    to climate impacts is critical for conservation planning and intervention strategies. A
    scenario-based population modelling approach was adopted with larval connectivity dynamics and
    environmental factors to assess coral reef resilience across the GBR using projected conditions under
    five Global Climate Models (GCMs). Projected coral
    cover was analysed for each reef's ability to maintain positive carbonate production budgets under
    future conditions, using coral cover as a proxy. Larval connectivity patterns did not correlate with
    increases in maintenance of positive carbonate budgets. Instead, reef depth emerged as the primary
    predictor, with deeper reefs benefitting from thermal refugia that reduce bleaching impacts. These
    findings suggest that depth is a tangible and pragmatic reef characteristic to consider in future
    intervention practices for coral reef restoration. These results have important implications for reef
    management, indicating that depth refugia should be prioritized for further consideration in
    conservation planning to maximize coral survival under continuing climate change.

keywords:
  - Coral reef
  - Climate change
  - Ecosystem resilience
  - Reef depth
  - Reef connectivity

filters:
  - _extensions/authors-block

date: last-modified

format:
  elsevier-docx:
    journal:
      name: Global Change Biology
      formatting: preprint
      cite-style: authoryear

bibliography: references.bib
csl: global-change-biology.csl

engine: julia
julia:
  exeflags: ["+1.11.5"]

crossref:
  custom:
  - kind: float
    reference-prefix: SFigure
    key: sfig
  - kind: float
    reference-prefix: STable
    key: stbl

editor:
  markdown:
    wrap: 92
    canonical: true
---
